---
title: "Humidity"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(DT)
library(terra)
library(rnaturalearth)
library(viridis)
library(rgeoboundaries)
```


## Downloading humidity data with `nasapower`

The `nasapower` package aims at making it quick and easy to automate downloading [NASA POWER](https://power.larc.nasa.gov/) (NASA Prediction of Worldwide Energy Resource) global meteorology, surface solar energy and climatology data.

Here, we will use the `nasapower` package to retrieve the relative humidity data for specific countries or for the world. 

We have also used the `nasapower` package to retrieve rainfall data [here](https://rspatialdata.github.io/rainfall).
The rainfall tutorial includes an introduction on the `nasapower` package and how its functions work. 

## Installing `nasapower` package 

We can install the package from CRAN and load it as follows.

```{r CRAN,eval=FALSE}
install.packages("nasapower")
```

```{r loading}
library(nasapower)
```

## Using `get_power()` to fetch data

First let us have a look at how to get the daily data for humidity in agriculture. This can be done using the `get_power()` function.

### Fetching daily data for single point

We use `get_power()` function arguments `pars = "RH2M"` which means relative Humidity at 2 meters, `temporal_average = "DAILY"`, and `longlat` equal to a single location.
   
```{r getdata}
data_RH <- get_power(community = "AG",
          lonlat = c(134.489563,-25.734968),
          dates = c("2010-09-23","2010-12-23"),
          temporal_average = "DAILY",
          pars = "RH2M")
data_RH %>% datatable(extensions = c('Scroller','FixedColumns'), options = list(
  deferRender = TRUE,
  scrollY = 350,
  scrollX = 350,
  dom = 't',
  scroller = TRUE,
  fixedColumns = list(leftColumns = 3)
))
```



### Fetching daily data for an area


```{r daily}
daily_humidity <- get_power(community = "AG",
          lonlat = c(150.5, -28.5 , 153.5, -25.5),
          pars = "RH2M",
          dates = c("2004-09-19","2004-09-29"),
          temporal_average = "DAILY")
daily_humidity %>% datatable(extensions = c('Scroller','FixedColumns'), options = list(
  deferRender = TRUE,
  scrollY = 350,
  scrollX = 350,
  dom = 't',
  scroller = TRUE,
  fixedColumns = list(leftColumns = 3)
))
```


### Fetching climatology data

Global data are obtained by setting `community = "AG"` and `temporal_average = "CLIMATOLOGY"`.

```{r climate,cache=TRUE}
climate_avg_RH <- get_power(community = "AG",
                         pars = "RH2M",
                         lonlat = "GLOBAL",
                         temporal_average = "CLIMATOLOGY"
)
climate_avg_RH %>% datatable(extensions = c('Scroller','FixedColumns'), options = list(
  deferRender = TRUE,
  scrollY = 350,
  scrollX = 350,
  dom = 't',
  scroller = TRUE,
  fixedColumns = list(leftColumns = 3)
))
```


## Creating spatial objects from `get_power()`

If you require spatial objects to work with, it is simple to convert the resultant data frame from `get_power()` to a spatial object using the `terra::rast(type = "...")`.

```{r climbox}
#getting the shapefiles for the world 
map <- ne_countries(returnclass = "sf")

climate_box <- split(climate_avg_RH,climate_avg_RH$PARAMETER)

climate_box <- lapply(climate_box, function(x){
  x["PARAMETER"] <- NULL
  x
})

climate_box <- lapply(X = climate_box, FUN= as.matrix)

#retrieving the humidity data using the above made climate_box function
relative_humid <- rast(climate_box$RH2M[,c(1:2,15)],
     crs = "+proj=eqc +lat_ts=0 +lat_0=0 +lon_0=0 +x_0=0 +y_0=0 +ellps=WGS84 +datum=WGS84 +units=m +no_defs", type = "xyz")

#converting above raster object into a dataframe for mapping 
RH_df <- as.data.frame(relative_humid, xy = TRUE, na.rm = TRUE)
rownames(RH_df) <- c()

#plotting the graph
ggplot() + 
  geom_raster(data = RH_df, aes(x = x, y = y, fill = ANN)) +
  geom_sf(data = map, inherit.aes = FALSE, fill = NA) + 
  scale_fill_viridis()+
  labs(title = "Relative Humidity",
       fill = "Humidity",
       subtitle = "Annual Relative Humidity at various parts of the world")
```


## Using `nasapower` with large geographical areas

We can plot the relative humidity data for large geographical areas using the shapefiles of different countries, raster cubes and finally plotting them together to show the humidity variation throughout the region. 

```{r sf}
library("rnaturalearth")
AUS <- ne_states(country = "Australia",
                 returnclass = "sf")
NT <- AUS[AUS$name == "Northern Territory",]
WA <- AUS[AUS$name == "Western Australia",]
```

```{r coord}
#creating raster objects
rasters <- rast(xmin = -180,
     xmax = 180,
     ymin = -90,
     ymax = 90,
     resolution = 0.5)
values(rasters) <- 1:ncell(rasters)
#retrieving only the coordinates from the NT & WA tables 
NT_coord <- crop(rasters,NT)
WA_coord <- crop(rasters,WA)
#getting the coordinated for the first administrative boundary
NT_coord <- mask(NT_coord,vect(NT))
#converting the coordinated into a dataframe with x and y values to map
NT_df <- as.data.frame(NT_coord, xy = TRUE, na.rm = TRUE)
rownames(NT_df) <- c()
#getting the US administrative boundaries shapefile, can alos use the ozmap package for australia shapefile
ggplot() + 
  geom_raster(data = NT_df, aes(x = x, y = y, fill = lyr.1))+ 
  geom_sf(data = AUS, inherit.aes = FALSE, fill = NA)+
  scale_fill_viridis()+
  theme_minimal()+
  labs(title = "Relative Humidity in Northern Territory,Australia", fill = "Humidity")
```

```{r wa-map}
WA_coord <- mask(WA_coord,vect(WA))
WA_df <- as.data.frame(WA_coord, xy = TRUE, na.rm = TRUE)
rownames(WA_df) <- c()
ggplot() + 
  geom_raster(data = WA_df, aes(x = x, y = y, fill = lyr.1))+ 
  geom_sf(data = AUS, inherit.aes = FALSE, fill = NA)+
  scale_fill_viridis()+
  theme_minimal()+
  labs(title = "Relative Humidity in Western Australia, Australia", fill = "Humidity")
```


Now let us have a look at the relative humidity monthly data for all the countries in the world: 

```{r month_world,out.height=750,out.width=800}
RH2M <- as.matrix(subset(climate_avg_RH, PARAMETER == "RH2M")[, -3])
RH2M_MONTH <- vector(mode = "list", length = 13)
names(RH2M_MONTH) <- colnames(RH2M)[-c(1:2,13)]
#converting each month into a spatRaster object using the for loop
for (k in colnames(RH2M)[-c(1:2,13)]) {
  RH2M_MONTH[[k]] <- rast(RH2M[, c("LON","LAT",k)],
                        crs = "+proj=eqc +lat_ts=0 +lat_0=0 +lon_0=0 +x_0=0 +y_0=0 +ellps=WGS84 +datum=WGS84 +units=m +no_defs",
                        type = "xyz")
  
}
RH2M <- c(rast(RH2M_MONTH))
plot(RH2M)
```

## References

- `nasapower` package: https://github.com/ropensci/nasapower

- NASAPOWER project: https://power.larc.nasa.gov/