---
title: "Humidity"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(DT)
library(terra)
library(rnaturalearth)
library(viridis)
library(rgeoboundaries)
```


## Downloading humidity data with `nasapower`

The `nasapower` package aims at making it quick and easy to automate downloading [NASA POWER](https://power.larc.nasa.gov/) (NASA Prediction of Worldwide Energy Resource) global meteorology, surface solar energy and climatology data.

Here, we will use the `nasapower` package to retrieve the relative humidity data for specific countries or for the world. 

We have also used the `nasapower` package to retrieve rainfall data [here](https://rspatialdata.github.io/rainfall).
The rainfall tutorial includes an introduction on the `nasapower` package and how its functions work. 

## Installing `nasapower` package 

We can install the package from CRAN and load it as follows.

```{r CRAN,eval=FALSE}
install.packages("nasapower")
```

```{r loading}
library(nasapower)
```

## Using `get_power()` to fetch data

First let us have a look at how to get the daily data for humidity in agriculture. This can be done using the `get_power()` function.

### Fetching daily data for single point

We use `get_power()` function arguments `pars = "RH2M"` which means relative umidity at 2 meters, `temporal_average = "DAILY"`, and `longlat` equal to a single location.
   
```{r getdata}
data_RH <- get_power(community = "AG",
          lonlat = c(134.489563,-25.734968),
          dates = c("2010-09-23","2010-12-23"),
          temporal_average = "DAILY",
          pars = "RH2M")

data_RH %>% datatable(extensions = c('Scroller','FixedColumns'), options = list(
  deferRender = TRUE,
  scrollY = 350,
  scrollX = 350,
  dom = 't',
  scroller = TRUE,
  fixedColumns = list(leftColumns = 3)
))
```



### Fetching daily data for an area


```{r daily}
daily_humidity <- get_power(community = "AG",
          lonlat = c(150.5, -28.5 , 153.5, -25.5),
          pars = "RH2M",
          dates = c("2004-09-19","2004-09-29"),
          temporal_average = "DAILY")

daily_humidity %>% datatable(extensions = c('Scroller','FixedColumns'), options = list(
  deferRender = TRUE,
  scrollY = 350,
  scrollX = 350,
  dom = 't',
  scroller = TRUE,
  fixedColumns = list(leftColumns = 3)
))
```


### Fetching climatology data

Global data are obtained by setting `community = "AG"` and `temporal_average = "CLIMATOLOGY"`.

```{r climate,cache=TRUE}
climate_avg_RH <- get_power(community = "AG",
                         pars = "RH2M",
                         lonlat = "GLOBAL",
                         temporal_average = "CLIMATOLOGY"
)
climate_avg_RH %>% datatable(extensions = c('Scroller','FixedColumns'), options = list(
  deferRender = TRUE,
  scrollY = 350,
  scrollX = 350,
  dom = 't',
  scroller = TRUE,
  fixedColumns = list(leftColumns = 3)
))
```



## Creating a map of annual humidity using all data retrieved


```{r}
library(rnaturalearth)
library(raster)

# Getting world map
map <- ne_countries(returnclass = "sf")

# Converting data to raster
r <- rasterFromXYZ(climate_avg_RH[, c("LON", "LAT", "ANN")])

# Converting the raster into a data.frame
r_df <- as.data.frame(r, xy = TRUE, na.rm = TRUE)

# Plot
ggplot() + 
  geom_raster(data = r_df, aes(x = x, y = y, fill = ANN)) +
  geom_sf(data = map, inherit.aes = FALSE, fill = NA) + 
  scale_fill_viridis()+
  labs(title = "Relative Humidity",
       fill = "Humidity",
       subtitle = "Annual Relative Humidity at various parts of the world")
```





## Creating a map of annual humidity using a subset of the data retrieved


```{r climbox}
library(rnaturalearth)

# Getting map for China
AUS <- ne_states(country = "Australia", returnclass = "sf")

# Getting administrative boundaries for regions
NT <- AUS[AUS$name == "Northern Territory", ]
WA <- AUS[AUS$name == "Western Australia", ]

# Converting data to raster
r <- rasterFromXYZ(climate_avg_RH[, c("LON", "LAT", "ANN")])

# Subset values for the region and converting the raster into a data.frame
rr <- mask(crop(r, NT), NT)
r_df <- as.data.frame(rr, xy = TRUE, na.rm = TRUE)

ggplot() + 
  geom_raster(data = r_df, aes(x = x, y = y, fill = ANN)) +
  geom_sf(data = AUS, inherit.aes = FALSE, fill = NA) + 
  scale_fill_viridis()+
  theme_minimal()+
  labs(title = "Relative Humidity in Northern Territory, Australia", fill = "Humidity")


# Subset values for the region and converting the raster into a data.frame
rr <- mask(crop(r, WA), WA)
r_df <- as.data.frame(rr, xy = TRUE, na.rm = TRUE)

ggplot() + 
  geom_raster(data = r_df, aes(x = x, y = y, fill = ANN)) +
  geom_sf(data = AUS, inherit.aes = FALSE, fill = NA) + 
  scale_fill_viridis()+
  theme_minimal()+
  labs(title = "Relative Humidity in Western Australia, Australia", fill = "Humidity")

```




## Creating maps of monthly humidity


```{r monthwise,out.width=800,out.height=750}
r <- list()
for(k in colnames(climate_avg_RH)[-c(1:3, 16)]) {
r[[k]] <- rasterFromXYZ(climate_avg_RH[, c("LON","LAT", k)])
}
r <- stack(r)
plot(r)
```


## References

- `nasapower` package: https://github.com/ropensci/nasapower

- NASAPOWER project: https://power.larc.nasa.gov/