---
title: "Population"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error=FALSE, message = FALSE, warning = FALSE)
```

# Downloading Population data using `wopr`

[`wopr`](https://github.com/wpgp/wopr) is an R package that provides access to the [WorldPop Open Population Repository](https://wopr.worldpop.org/) and gets estimates of population sizes for specific demographic groups.

## Installing `wopr`

The `wopr` package can be directly downloaded from the GitHub repository of the package. For this we use the `remotes` package which allows easy installation of R packages from remote repositories such as GitHub. So we install and load the `remotes` package and use it to install the `wopr` package as follows

```{r install-wopr, eval=FALSE}
install.packages("remotes")
library(remotes)
remotes::install_github('wpgp/wopr')
```

## Downloading data

To download the population data we first retrieve the WOPR data catalogue and then subset the data frame for the country we are interested in. In this tutorial we use 'NGA' which refers to Nigeria. WOPR uses [ISO country codes](https://en.wikipedia.org/wiki/ISO_3166-1_alpha-3) to abbreviate country names.

Then we simply download the data for the subsetted catalogue selection using the `downloadData()` function. Note that `downloadData()` will not download files larger than 100MB by default unless the `maxsize` argument is changed (see `?downloadData`)

```{r, download-pop-data, eval=FALSE}
library(wopr)
# Retrieve the WOPR data catalogue
catalogue <- getCatalogue()

# Select files from the catalogue by subsetting the data frame
selection <- subset(catalogue,
                    country == 'NGA' &
                      category == 'Population' & 
                      version == 'v1.2')

# Download selected files
downloadData(selection)
```

## Understanding the downloaded data

The download will create a folder named `./wopr` in the R working directory for the downloaded files and a spreadsheet with information on the downloaded files will be available at `./wopr/wopr_catalogue.csv`

Since we downloaded the population estimates of Nigeria(NGA), zipped data files will be available at `./wopr/NGA/population/v1.2` which will then have to be manually unzipped.

The folder structure will look as follows

```
working directory
└── wopr
    ├── wopr_catalogue.csv
    └── NGA
        └── population
            └── v1.2
                ├── NGA_population_v1_2_admin.zip
                ├── NGA_population_v1_2_gridded.zip
                ├── NGA_population_v1_2_mastergrid.tif
                ├── NGA_population_v1_2_methods.zip
                └── NGA_population_v1_2_README.pdf
```

Since the zipped file `./wopr/NGA/population/v1.2/NGA_population_v1_2_admin.zip` contains population totals for administrative units in Nigeria (i.e. states and local government areas) and shapefiles for the administrative boundaries, we unzip this file and place it in the same directory. The folder structure will now look as follows

```
working directory
└── wopr
    ├── wopr_catalogue.csv
    └── NGA
        └── population
            └── v1.2
                ├── NGA_population_v1_2_admin.zip
                ├── ...
                └── NGA_population_v1_2_admin
                  ├── NGA_population_v1_2_admin_level0.csv
                  ├── ...
```

The unzipped folder (at `./wopr/NGA/population/v1.2/NGA_population_v1_2_admin`) will contain multiple `.csv` files and `.shp` files containing data on multiple administrative levels of Nigeria, which we will use to visualise the population.

The `.csv` files contain a summary of the distributions of the population estimates, while the `.shp` files contain vector geospatial data for the respective administrative divisions.

## Population of different administrative levels of Nigeria

In this tutorial we use the `rgdal` package to read in the downloaded shapefiles into R. The `rgdal` package provides bindings for the [Geospatial Data Abstraction Library](https://gdal.org/) allowing us to read in raster and vector geospatial data formats.

We will also use the `here` package to get file paths to the working directory so that we can simply use relative paths to import files.

First we read in the administrative level 2 data of Nigeria as follows

```{r read-admlvl2-data}
# install.packages(c("rgdal", "here"))
library(rgdal)
library(here)

admin_level2_shape <- readOGR(dsn=here::here("wopr/NGA/population/v1.2/NGA_population_v1_2_admin/NGA_population_v1_2_admin_level2_boundaries.shp"), verbose=FALSE)
```

The `ggplot2` package can be used to plot the administrative boundaries and visualise the population. `ggplot2` allows us to easily visualise simple feature objects using the `geom_sf()` function, but we will need to convert the shapefile we have into a simple feature object. This can be done using the `st_as_sf()` function in the `sf` package.

```{r vis-admlvl2-data}
# install.packages(c("sf", "ggplot2"))
library(sf)
library(ggplot2)
admin_level2_shape %>% 
  sf::st_as_sf() %>% 
  ggplot(aes(fill = mean)) +
  geom_sf()
```


Similarly the administrative level 3 data can also be visualised as follows


```{r vis-admlvl3-data}
admin_level3_shape <- readOGR(here::here("wopr/NGA/population/v1.2/NGA_population_v1_2_admin/NGA_population_v1_2_admin_level3_boundaries.shp"), verbose=FALSE)

admin_level3_shape %>% 
  sf::st_as_sf() %>% 
  ggplot(aes(fill = mean)) +
  geom_sf()
```

## References

- `wopr` repository: https://github.com/wpgp/wopr  
- `ggplot2` package: https://ggplot2.tidyverse.org/