---
title: "Rainfall"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(DT)
library(terra)
```

# `nasapower` package
 
This package aims at making it quick and easy to automate downloading NASA [POWER](https://power.larc.nasa.gov/)(Prediction of Worldwide Energy Resource) global meteorology, surface solar energy and climatology data in your R session as a tidy data frame for analysis and use in modeling or other purposes using `get_power()`. 


## Installing `nasapower` package 

You can install the package from either CRAN using:

```{r CRAN,eval=FALSE}
install.packages("nasapower")
```

or by using the devtools package directly from github using the following code:

```{r git, eval=FALSE}
if (!require(devtools)) {
  install.packages("devtools")
}

devtools::install_github("ropensci/nasapower")
```

```{r loading}
library(nasapower)
```

## Using `get_power()` to fetch data

The `get_power()` function has five arguments and returns a data frame with a metadata header in the current R session.
It has the following arguments:

- **Community**: It can be passed with either "AG", "SB", or "SSE".

    *AG*: provides access to the agroclimatology archive, which contains industry-friendly parameters for input to crop models.
    
    *SB*: provides access to the sustainable buildings archive, which contains parameters for the building community.
    
    *SSE*: provides access to the renewable energy archive, which contains parameters very specific to assist in the design of solar and wind powered renewable energy systems.
    
- **temporal_average**: supported values are "DAILY", "INTERANNUAL", "CLIMATOLOGY".
  
    *DAILY* : the daily average of `pars` by day, month and year.
    
    *INTERANNUAL*: the monthly average of `pars` by year.
    
    *CLIMATOLOGY*: the monthly average of `pars` at the surface of the earth for a given month averaged for that month over the 30 year period.
    
- **lonlat**: 

    *For Single point*: supply a length-two numeric vector giving the decimal degree longitude an dlatitude in that order for the data to download. 
    
    *For regional coverage*: supply a length-four numeric as lower left(lon,lat) and upper right(lon,lat) coordinates as `lonlat = c(xmin,ymin,ymax,ymax)`
    
    *For global coverage*: to get globval coverage for CLIMATOLOGY, supply "GLOBAL" while also specifying "CLIMATOLOGY" for the argument `temporla_average`.
    
- **dates**: If only one date is provided, it will be treated as both the start and the end date and only a day's values will be returned. 

    When the `temporal_average` is set to "INTERANNUAL", use only two year values, eg, dates = c(1983,2010). 

    This argument should not be used when `temporal_average` is set to "CLIMATOLOGY".

To know the different weather values from POWER provided within this function type `?get_power`, and in the arguments section, click on the highlighted parameters, which goes to a page which has all the available parameters. For rainfall, we use the "PRECTOT". 

### Fetching daily data for single point:
   
```{r getdata}
data <- get_power(community = "SSE",
          lonlat = c(134.489563,-25.734968),
          dates = c("2000-01-01","2000-05-01"),
          temporal_average = "DAILY",
          pars = "PRECTOT")
data %>% datatable(extensions = c('Scroller','FixedColumns'), options = list(
  deferRender = TRUE,
  scrollY = 350,
  scrollX = 350,
  dom = 't',
  scroller = TRUE,
  fixedColumns = list(leftColumns = 3)
))
```

### Fetching daily data for an area:

```{r}
daily_area <- get_power(community = "AG",
          lonlat = c(150.5, -28.5 , 153.5, -25.5),
          pars = "PRECTOT",
          dates = c("2004-09-19","2004-09-29"),
          temporal_average = "DAILY")
daily_area %>% datatable(extensions = c('Scroller','FixedColumns'), options = list(
  deferRender = TRUE,
  scrollY = 350,
  scrollX = 350,
  dom = 't',
  scroller = TRUE,
  fixedColumns = list(leftColumns = 3)
))
```

### Fetching climatology data:

global data are only available for the climatology `temporal_average` like we discussed earlier, setting these arguments as such will fetch the global values.

```{r climate}
climate_avg <- get_power(community = "AG",
                         pars = "PRECTOT",
                         lonlat = "GLOBAL",
                         temporal_average = "CLIMATOLOGY"
)

climate_avg %>% datatable(extensions = c('Scroller','FixedColumns'), options = list(
  deferRender = TRUE,
  scrollY = 350,
  scrollX = 350,
  dom = 't',
  scroller = TRUE,
  fixedColumns = list(leftColumns = 3)
))
```


## Creating spatial objects from `get_power()`

If you require spatial objects to work with, it is simple to convert the resultant data frame from `get_power()` to a spatial object using the `terra::rast(type = "...")`


```{r climbox}
climate_box <- split(climate_avg,climate_avg$PARAMETER)

climate_box <- lapply(climate_box, function(x){
  x["PARAMETER"] <- NULL
  x
})

climate_box <- lapply(X = climate_box, FUN= as.matrix)

PRECTOT <- rast(climate_box$PRECTOT[,c(1:2,14)],
     crs = "+proj=eqc +lat_ts=0 +lat_0=0 +lon_0=0 +x_0=0 +y_0=0 +ellps=WGS84 +datum=WGS84 +units=m +no_defs",
                 type = "xyz")

plot(PRECTOT, plg = list(title = "World rainfall extent"))
```

## Using `nasapower` with large geographical areas

```{r eval=FALSE}
library("rnaturalearth")

MEX <- ne_states(country = "Mexico",
                 returnclass = "sf")

CH <- MEX[MEX$name == "Chihuahua",]
DU <- MEX[MEX$name == "Durango",]
```

```{r eval=FALSE}
library("ggplot2")
library("purrr")
r <- rast(xmin = -180,
     xmax = 180,
     ymin = -90,
     ymax = 90,
     resolution = 0.5)

values(r) <- 1:ncell(r)

CH_coords <- crop(r,CH)
DU_coords <- crop(r,DU)

CH_coords <- mask(CH_coords,sf:::as_Spatial(CH))
plot(CH_coords)
```




