---
title: "Rainfall"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(DT)
library(terra)
library(rnaturalearth)
library(viridis)
library(rgeoboundaries)
```

## Downloading rainfall data with `nasapower`
 
The `nasapower` package aims at making it quick and easy to automate downloading [NASA POWER](https://power.larc.nasa.gov/) (NASA Prediction of Worldwide Energy Resource) global meteorology, surface solar energy and climatology data.


## Installing `nasapower` package 

We can install the package from either CRAN using

```{r CRAN,eval=FALSE}
install.packages("nasapower")
```

or by using the `devtools` package directly from GitHub using the following code:

```{r git, eval=FALSE}
if (!require(devtools)) {
  install.packages("devtools")
}
devtools::install_github("ropensci/nasapower")
```

```{r loading}
library(nasapower)
```

## Using `get_power()` to fetch data

The `get_power()` function has five arguments and returns a data frame with a metadata header in the current R session.
It has the following arguments:

- `pars`: character vector of solar, meteorological or climatology parameters to download.

- `community`: character vector providing community name. Supported values are `"AG"`, `"SB"` or `"SSE"`.

`"AG"`: provides access to the agroclimatology archive, which contains industry-friendly parameters for input to crop models.
    
`"SB"`: provides access to the sustainable buildings archive, which contains parameters for the building community.
    
`"SSE"`: provides access to the renewable energy archive, which contains parameters very specific to assist in the design of solar and wind powered renewable energy systems.
    
- `temporal_average`: temporal average for data being queried. Supported values are `"DAILY"`, `"INTERANNUAL"` or `"CLIMATOLOGY"`.
  
`"DAILY"`: daily average of `pars` by day, month and year.
    
`"INTERANNUAL"`: monthly average of `pars` by year.
    
`"CLIMATOLOGY"`: monthly average of `pars` at the surface of the earth for a given month averaged for that month over the 30 year period.
    
- `lonlat`: numeric vector of geographic coordinates for a cell or region or `"GLOBAL"` for global coverage.

For single point supply a length-two numeric vector giving the decimal degree longitude an dlatitude in that order for the data to download. 
    
For regional coverage supply a length-four numeric as lower left (lon, lat) and upper right (lon, lat) coordinates as `lonlat = c(xmin, ymin, ymax, ymax)`
    
To get global coverage for `CLIMATOLOGY`, supply `GLOBAL` while also specifying `CLIMATOLOGY` for the argument `temporal_average`.
    
- `dates`: start and end dates. If only one date is provided, it will be treated as both the start and the end date and only a day's values will be returned. 
When the `temporal_average` is set to `INTERANNUAL`, use only two year values, e.g., `dates = c(1983, 2010)`. 
This argument should not be used when `temporal_average` is set to `"CLIMATOLOGY"`.

To know the different weather values from POWER provided within this function type `?get_power`, and in the arguments section, click on the highlighted parameters, which goes to a page which has all the available parameters.

To download rainfall, we use the `pars = "PRECTOT"`. 

### Fetching daily data for single point
   
```{r getdata}
data <- get_power(community = "SSE",
          lonlat = c(134.489563,-25.734968),
          dates = c("2000-01-01","2000-05-01"),
          temporal_average = "DAILY",
          pars = "PRECTOT")

data %>% datatable(extensions = c('Scroller','FixedColumns'), options = list(
  deferRender = TRUE,
  scrollY = 350,
  scrollX = 350,
  dom = 't',
  scroller = TRUE,
  fixedColumns = list(leftColumns = 3)
))
```

### Fetching daily data for an area

```{r daily}
daily_area <- get_power(community = "AG",
          lonlat = c(150.5, -28.5 , 153.5, -25.5),
          pars = "PRECTOT",
          dates = c("2004-09-19","2004-09-29"),
          temporal_average = "DAILY")

daily_area %>% datatable(extensions = c('Scroller','FixedColumns'), options = list(
  deferRender = TRUE,
  scrollY = 350,
  scrollX = 350,
  dom = 't',
  scroller = TRUE,
  fixedColumns = list(leftColumns = 3)
))
```

### Fetching climatology data

Global data are only available for the climatology `temporal_average` like we discussed earlier, setting these arguments as such will fetch the global values.

```{r climate}
climate_avg <- get_power(community = "AG",
                         pars = "PRECTOT",
                         lonlat = "GLOBAL",
                         temporal_average = "CLIMATOLOGY"
)

climate_avg %>% datatable(extensions = c('Scroller','FixedColumns'), options = list(
  deferRender = TRUE,
  scrollY = 350,
  scrollX = 350,
  dom = 't',
  scroller = TRUE,
  fixedColumns = list(leftColumns = 3)
))
```


## Creating a map of annual rainfall using all data retrieved


```{r}
library(rnaturalearth)
library(raster)

# Getting world map
map <- ne_countries(returnclass = "sf")

# Converting data to raster
r <- rasterFromXYZ(climate_avg[, c("LON", "LAT", "ANN")])

# Converting the raster into a data.frame
r_df <- as.data.frame(r, xy = TRUE, na.rm = TRUE)

# Plot
ggplot() + 
  geom_raster(data = r_df, aes(x = x, y = y, fill = ANN)) +
  geom_sf(data = map, inherit.aes = FALSE, fill = NA) + 
  scale_fill_viridis()+
  labs(title = "Rainfall in inches",
       fill = "Annual Rainfall",
       subtitle = "Annual rainfall at various parts of the world")
```


## Creating a map of annual rainfall using a subset of the data retrieved


```{r climbox}
library(rnaturalearth)

# Getting map for China
CHI <- ne_states(country = "China", returnclass = "sf")

# Getting administrative boundaries for regions
GA <- CHI[CHI$name == "Gansu", ]
XI <- CHI[CHI$name == "Xinjiang", ]

# Converting data to raster
r <- rasterFromXYZ(climate_avg[, c("LON", "LAT", "ANN")])

# Subset values for the region and converting the raster into a data.frame
rr <- mask(crop(r, GA), GA)
r_df <- as.data.frame(rr, xy = TRUE, na.rm = TRUE)

ggplot() + 
  geom_raster(data = r_df, aes(x = x, y = y, fill = ANN)) +
  geom_sf(data = CHI, inherit.aes = FALSE, fill = NA) + 
  scale_fill_viridis()+
  theme_minimal()+
  labs(title = "Rainfall in Gansu, China", fill = "Rainfall")


# Subset values for the region and converting the raster into a data.frame
rr <- mask(crop(r, XI), XI)
r_df <- as.data.frame(rr, xy = TRUE, na.rm = TRUE)

ggplot() + 
  geom_raster(data = r_df, aes(x = x, y = y, fill = ANN)) +
  geom_sf(data = CHI, inherit.aes = FALSE, fill = NA) + 
  scale_fill_viridis()+
  theme_minimal()+
  labs(title = "Rainfall in Xinjiang, China", fill = "Rainfall")

```


## Creating maps of monthly rainfall


```{r monthwise,out.width=800,out.height=750}
r <- list()
for(k in colnames(climate_avg)[-c(1:3, 16)]) {
r[[k]] <- rasterFromXYZ(climate_avg[, c("LON","LAT", k)])
}
r <- stack(r)
plot(r)
```

## References

- `nasapower` package: https://github.com/ropensci/nasapower

- NASAPOWER project: https://power.larc.nasa.gov/
